Supplementary Materials- A Bayesian approach to burial networks to infer social changes in northeastern Taiwan during the European colonization period

### 2021-01-27 analysis/paper/paper.Rmd
merge changes in highlights
edit ERGMs background and result sections in paper rmd, fix a typo in SI

---
title: "A Bayesian approach to burial networks to infer social changes in northeastern Taiwan during the European colonization period"
author:
  - Li-Ying Wang
  - Ben Marwick
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Burials provide valuable information to study social structures based on the assumption that burials and associated grave goods can represent social roles and relations in a society. To study social relationships, network analysis has been increasingly applied to archaeological data to explore network structures and patterns that can infer interactions and relationship between entities. Different from traditional static networks, statistical approaches to network analysis, such as exponential random graph models (ERGMs), provides a way to test hypotheses on dynamic process of network formation. However, the computational difficulty and sensitivity to uncertainties limits the application of ERGMs. In this paper, we introduce a Bayesian framework on ERGMs that enables an efficient computational process, effective quantification for uncertainty, and robust model evaluation of network properties. We tested the hypothesis of social changes relative to the arrival of Europeans by studying burial data from Kiwulan, an Iron Age site in northeastern Taiwan. The results indicate a transition from ties based on ritual objects to wealth objects, and a more centralized structure with social differentiation after European presence in the 17th century. Our case study demonstrates the effectiveness of Bayesian network analysis for archaeological data, and expands the use of burials in understanding the impacts of colonial presence on Indigenous groups in a pericolonial context.
keywords: |
  Bayesian modeling; network analysis; Monte-carlo simulation; Burials; European colonialism; social differentiation; Taiwan
highlights: |
  1. We present Bayesian inference on exponential random graph models (ERGMs) to model archaeological networks.
  2. We test social processes by simulating two sets of burial data from different time periods.
  3. Differences in network statistics imply change over time in social structures influenced by foreign presence. 
  4. Bayesian ERGMs are efficient and useful for evaluating and interpreting networks. 
---

@@

<!--fit in a bigger historical picture in a periclonial context: move to SI -->

@@

<!-- Bayesian inference for Exponential Random Graph Models --> 
Bayesian approaches to Exponential Random Graph Models (ERGMs) are effective tools for network modeling by incorporating prior information about the network configurations to better understand dependencies of network variables and improve computational issues in ERGMs [@Caimo2017; @Lehmann2020]. The advantage that Bayesian modeling has over traditional ERGMs is the application of Markov chain Monte Carlo (MCMC) simulation using the approximate exchange algorithm [@Caimo2011]. MCMC avoids doubly-intractable computations by directly sampling from the not normalized part of the posterior, which alleviates the computational problems and gives a better convergence results. This enable us to deal with complicated dependence patterns with ease, providing better estimations for complex social network models with heterogeneous data [@Caimo2017; @Snijders2006]. After fitting an ERMG with an MCMC algorithm, a Bayesian approach generates posterior probabilities that incorporates our sample data and prior information derived from previous data or our assumptions through summary statistics from estimates of the ERGM parameters [@Caimo2015; @Nemmers2019]. This also allows uncertainty quantification. Posterior probability estimates the effect of ERGM parameters by looking at the posterior mean and 95% credible intervals, which can replace p-values for assessing a null hypothesis [@Caimo2017bayesian]. The typical criteria for interpreting the posteriors is that an odds ratio greater than one means a positive effect of a parameter, while odds ratios less than one represents negative effect. In addition, Bayesian approaches are also useful to deal with missing data, which is often a problem leading to misinterpretation of networks, especially for archaeological studies. @Koskinen2010 shows that the effect of missing data can be reduced with Bayesian modeling that can predict, on the average, 80% of the ties when a third of data missing.

@@

<!--Model fitting: move to SI -->

@@

<!--posterior stats for pre-E-->
We examine the estimates from the posterior distributions to compare their differences in structure of the simulated networks (Table \@ref(tab:table-posterior-stats); Figure \@ref(fig:figure-posterior-distribution)). For the pre-European network simulations, the nodal covariates, ritual-homophily, have significant effects on the formation of relations between burials. The positive effect demonstrates that burials with ritual pottery tend to form relations. Despite positive values for some covariates, such as age-homophily, sex-homophily and burial value-homophily, they do not show a significant tendency to form relations duo to value of zero in the confidence intervals. Similarly, the dyadic covariate, physical distance, shows no significant effect, indicating that physical proximity between burials does not affect the formation of relations. For the endogenous network effects, transitivity (gwesp) present significant positive effect, while popularity (gwdegree) demonstrates negative effect. The high positive value for transitivity suggests a tendency of burials to be clustered in closed transitive structures, indicative of the presence of multiple corporate groups sharing burial goods in common. The negative popularity shows there is a tendency toward decentralization that reflects most burials have similar number of ties without any prominent ones. This might imply that individuals have equal access to trade goods in terms of the flow of goods.

@@

<!--posterior stats for for post-E and their comparison-->
For the network after the arrival of Europeans, the nodal covariates of burial value-homophily show positive effects, while the ritual-homophily and sex-homophily have a negative effect. Despite the effect is not strong, it could indicate the burials in the same wealth level tend to form relations. On the contrary, the same sex burials and burial with ritual pottery in common tend not to form relations. There are no significant effects for age-homophily and the dyadic covariate, physical distance. Similar to the pre-European network, the endogenous network variable, transitivity (gwesp) demonstrates a significant positive effect but much lower than the effect of pre-European network. In contrast, the popularity (gwdegree) shows a significantly higher positive effect than the effect of pre-European network. This means there is a tendency toward centralization that reflects a limited number of burials have many more ties than others. This implies that a few central individuals may have better access to trade goods and developed personal networks through wealth accumulation. In general, the post-European network has a smaller transitivity effect and a positive popularity effect than the pre-European network. This may suggest a reduced tendency toward clustering but high tendency toward centralization after the European presence. Both posterior estimates present symmetric distributions, with the posterior means close to posterior medians (Table \@ref(tab:table-posterior-stats)).

>>>>

---
title: "A Bayesian approach to burial networks to infer social changes in northeastern Taiwan during the European colonization period"
author:
  - Li-Ying Wang
  - Ben Marwick
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Burials provide valuable information to study social structures based on the assumption that burials and associated grave goods can represent social roles and relations in a society. To study social relationships, network analysis has been increasingly applied to archaeological data to explore network structures and patterns that can infer interactions and relationship between entities. Different from traditional static networks, statistical approaches to network analysis, such as exponential random graph models (ERGMs), provides a way to test hypotheses on dynamic process of network formation. However, the computational difficulty and sensitivity to uncertainties limits the application of ERGMs. In this paper, we introduce a Bayesian framework on ERGMs that enables an efficient computational process, effective quantification for uncertainty, and robust model evaluation of network properties. We tested the hypothesis of social changes relative to the arrival of Europeans by studying burial data from Kiwulan, an Iron Age site in northeastern Taiwan. The results indicate a transition from ties based on ritual objects to wealth objects, and a more centralized structure with social differentiation after European presence in the 17th century. Our case study demonstrates the effectiveness of Bayesian network analysis for archaeological data, and expands the use of burials in understanding the impacts of colonial presence on Indigenous groups in a pericolonial context.
keywords: |
  Bayesian modeling; network analysis; Monte-carlo simulation; Burials; European colonialism; social differentiation; Taiwan
highlights: |
  1. We present Bayesian inference on exponential random graph models (ERGMs) to model archaeological networks.
  2. We explore social processes and network structures in archaeological data from from two time periods.
  3. Differences in network statistics imply change over time in social structures influenced by foreign presence. 
  4. Bayesian ERGMs are efficient and useful for evaluating and interpreting networks. 
---

@@

@@

<!-- Bayesian inference for Exponential Random Graph Models --> 
Bayesian approaches to Exponential Random Graph Models (ERGMs) are effective tools for network modeling by incorporating prior information about the network configurations to better understand dependencies of network variables and improve computational issues in ERGMs [@Caimo2017; @Lehmann2020]. Prior information is derived from previous data or assumptions of our data. The advantage that Bayesian modeling has over traditional ERGMs is the application of Markov chain Monte Carlo (MCMC) simulation using the approximate exchange algorithm [@Caimo2011]. Bayesian ERGMs avoid doubly-intractable computations by directly sampling from the not normalized part of the posterior, which alleviates the computational problems and gives a better convergence results. This enable us to deal with complicated dependence patterns with ease, providing better estimations for complex social network models with heterogeneous data [@Caimo2017; @Snijders2006]. After fitting an ERMG with an MCMC algorithm, the Bayesian approach generates posterior probabilities incorporating our sample data and prior information that estimates the effect of each ERGM parameter in our models [@Caimo2015; @Nemmers2019]. This allows uncertainty quantification by looking at the posterior mean and 95% credible intervals, or the odds ratios (conditional probability of observing an edge) that can replace p-values for assessing a null hypothesis [@Caimo2017bayesian]. The typical criteria for interpreting an significant effect of a parameter is that its credible interval does not contain zero, or an odds ratio is greater/less than one, meaning a significant positive/negative effect. In addition, Bayesian approaches are also useful to deal with missing data, which is often a problem leading to misinterpretation of networks, especially for archaeological studies. @Koskinen2010 shows that the effect of missing data can be reduced with Bayesian modeling that can predict, on the average, 80% of the ties when a third of data missing.

@@

@@

<!--posterior stats for pre-E-->
We examine the estimates from the posterior distributions to compare their differences in structure of the simulated networks (Table \@ref(tab:table-posterior-stats); Figure \@ref(fig:figure-posterior-distribution)). For nodal covariates in the pre-European model, ritual-homophily has a significant effect on the formation of relations between burials, while burial value-homophily has a negative effect. This demonstrates that burials with ritual pottery tend to form relations, but burials in the same value level tend not to. Despite positive mean value for some covariates, such as age-homophily, sex-homophily, they do not show a significant tendency duo to value of zero in the confidence intervals. Similarly, the dyadic covariate, physical distance, shows no significant effect, indicating that physical proximity between burials does not affect the formation of relations. For the endogenous network effects, transitivity (gwesp) present significant positive effect, while popularity (gwdegree) demonstrates negative effect. The high positive value for transitivity suggests a tendency of burials to be clustered in closed transitive structures, indicative of the presence of multiple corporate groups sharing burial goods in common. In contrast, the strong negative popularity shows there is a tendency toward decentralization that reflects most burials have similar number of ties without any prominent ones. This might imply that individuals have equal access to trade goods in terms of the flow of goods.

@@

<!--posterior stats for for post-E and their comparison-->
For the network model after the arrival of Europeans, the nodal covariates of burial value-homophily show significant positive effects, while the ritual-homophily and sex-homophily have a negative effect. This indicates the burials in the same wealth level tend to form relations. On the contrary, the same sex burials and burial with ritual pottery in common tend not to form relations. There are no significant effects for age-homophily and the dyadic covariate, physical distance. Similar to the pre-European network, the endogenous network variable, transitivity (gwesp) demonstrates a significant positive effect but much lower than the effect of pre-European network. In contrast, the popularity (gwdegree) shows a significantly higher positive effect than the effect of pre-European network. This means there is a tendency toward centralization that reflects a limited number of burials have many more ties than others. This implies that a few central individuals might have better access to trade goods and developed personal networks through wealth accumulation. In general, the post-European network model has a smaller transitivity effect and a positive popularity effect than the pre-European network model. This may suggest a reduced tendency toward clustering but high tendency toward centralization after the European presence. In addtion, both posterior estimates present symmetric distributions, with the posterior means close to posterior medians (Table \@ref(tab:table-posterior-stats)).

### 2021-01-27 analysis/paper/supplementary-material.Rmd
edit ERGMs background and result sections in paper rmd, fix a typo in SI

## Kiwulan in the 17th century northeastern Taiwan {.unnumbered}

<!--fit in a bigger historical picture in a periclonial context: move to SI -->
The involvement in long-distance exchange is often associated with changes in political-economic strategies to a network mode, where ambitious individuals are able to build their local power base and personal influence through the distribution of high-value goods [@Blanton1996; @Feinman2000; @Klehm2017]. Northeastern Taiwan was involved in a regional exchange in East Asia since the 14th century and long-distance trade during the European presence in the 17th century [@Chen2005; @Wang2007]. In 1626, the Spanish built their forts in Heping dao and Tamsui, northern Taiwan, and later were defeated and taken over by the Dutch in 1642 [@Andrade2007]. Indigenous people in northeastern Taiwan first encountered colonial power in 1632 when they were attacked by the Spanish and later in 1647 by the Dutch, who also asked them to pay annual tributes in a flexible way [@Borao2009; @Kang2016]. The Dutch would give local Indigenous leaders symbolic items or trade goods to assert their political authority and consolidate their relationship with Indigenous societies.

>>>>

# Kiwulan in the 17th northstern Taiwan in a pericolonial context {-}

<!--fit in a bigger historical picture in a pericolonial context: move to SI -->
The involvement in long-distance exchange is often associated with changes in political-economic strategies to a network mode, where ambitious individuals are able to build their local power base and personal influence through the distribution of high-value goods [@Blanton1996; @Feinman2000; @Klehm2017]. Northeastern Taiwan was involved in a regional exchange in East Asia since the 14th century and long-distance trade during the European presence in the 17th century [@Chen2005; @Wang2007]. In 1626, the Spanish built their forts in Heping dao and Tamsui, northern Taiwan, and later were defeated and taken over by the Dutch in 1642 [@Andrade2007]. Indigenous people in northeastern Taiwan first encountered colonial power in 1632 when they were attacked by the Spanish and later in 1647 by the Dutch, who also asked them to pay annual tributes in a flexible way [@Borao2009; @Kang2016]. The Dutch would give local Indigenous leaders symbolic items or trade goods to assert their political authority and consolidate their relationship with Indigenous societies.
